---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)

# set plot colors
palette("Tableau 10")
palette(adjustcolor(palette(), alpha.f = 0.8))
```

# opa

<!-- badges: start -->
![](https://www.r-pkg.org/badges/version-ago/opa?color=orange)
![](https://cranlogs.r-pkg.org/badges/grand-total/opa) 
<!-- badges: end -->

An R package for ordinal pattern analysis.

## Installation

opa can be installed from CRAN with:

```{r install_opa, eval=FALSE}
install.packages("opa")
```

You can install the development version of opa from [GitHub](https://github.com/) with:

```{r install_github, eval=FALSE}
# install.packages("remotes")
remotes::install_github("timbeechey/opa")
```

## Citation

To cite opa in your work you can use the output of:

```{r citation, eval=FALSE, message=FALSE, warning=FALSE}
citation(package = "opa")
```

## Background

`opa` is an implementation of methods described in publications including [Thorngate (1987)](https://doi.org/10.1016/S0166-4115(08)60083-7) and [Grice et al. (2015)](https://doi.org/10.1177/2158244015604192). Thorngate (1987) attributes the original idea to:

Parsons, D. (1975). _The directory of tunes and musical themes_. S. Brown.

## How ordinal pattern analysis works

Ordinal pattern analysis is similar to Kendall's Tau. Whereas Kendall's tau is a measure of similarity between two data sets in terms of rank ordering, ordinal pattern analysis is intended to quantify the match between a hypothesis and patterns of individual-level data across conditions or measurement instances.

Ordinal pattern analysis works by comparing the relative ordering of pairs of observations and computing whether those pairwise relations are matched by a hypothesis. Each pairwise ordered relation is classified as an increases, a decrease, or as no change. These classifications are encoded as 1, -1 and 0, respectively. For example, a hypothesis of a monotonic increase in a response variable across four experimental conditions can be specified as:

```{r hypothesis1}
h <- c(1, 2, 3, 4)
```

Note that the absolute values are not important, only their relative ordering. The hypothesis `h` encodes six pairwise relations, all increases: `1 1 1 1 1 1`.

A row of individual data representing measurements across four conditions, such as:

```{r vector1}
dat <- c(65.3, 68.8, 67.0, 73.1)
```

encodes six ordered pairwise relations `1 1 1 -1 1 1`. The percentage of orderings which are correctly classified by the hypothesis (_PCC_) is the main quantity of interest in ordinal pattern analysis. Comparing `h` and `dat`, the PCC is `5/6 = 0.833` or 83.3%. A hypothesis which generates a greater PCC is preferred over a hypothesis which generates a lower PCC for given data.

It is also possible to calculate a _chance-value_ for a PCC which is equal to the chance that a PCC at least as great as the PCC of the observed data could occur as a result of a random re-ordering of the data. Chance values can be computed using either a permutation test or a randomization test.

## Using `opa`

```{r load_opa}
library(opa)
```

A hypothesized relative ordering of a response variable across conditions is specified with a numeric vector:

```{r hypothesis2}
h <- c(1, 2, 4, 3)
```

The hypothesis can be plotted with the `plot_hypothesis()` function:

```{r plot_hypothesis, fig.width=4, fig.height=4, fig.align="center"}
plot_hypothesis(h)
```

Data should be in _wide_ format with one column per measurement condition and one row per individual:

```{r load_data}
set.seed(123)

dat <- data.frame(t1 = rnorm(20, mean = 12, sd = 2),
                  t2 = rnorm(20, mean = 15, sd = 2),
                  t3 = rnorm(20, mean = 20, sd = 2),
                  t4 = rnorm(20, mean = 17, sd = 2))
                  
round(dat, 2)
```

An ordinal pattern analysis model to consider how the hypothesis `h` matches each individual pattern of results in `dat` can be fitted using:

```{r opamod1}
opamod <- opa(dat, h, cval_method = "exact")
```

Specifying `cval_method = "exact"` computes chance values using a permutation test which considers every possible ordering of the data to determine the exact chance of a PCC at least as high as the PCC of the observed data. Specifying instead `cval_method = "stochastic"` computes chance values based on PCCs computed from a set of random orderings of the data. The number of random orderings to consider can be specified with the `nreps` argument which defaults to `1000L`.

A summary of the model output can be viewed using:

```{r summary1}
summary(opamod)
```

Individual-level model output can be plotted using:

```{r plot_opamod1, fig.width=9, fig.height=4.5, fig.align="center"}
plot(opamod)
```

To aid in visual interpretation, individual PCCs and c-values can also be plotted relative to user-specified thresholds:

```{r threshold_plots, fig.width=5, fig.height=5, fig.align="center", fig.show="hold", out.width="50%"}
pcc_plot(opamod, threshold = 90)
cval_plot(opamod, threshold = 0.1)
```

### Pairwise comparison of measurement conditions

Pairwise comparisons of measurement conditions can be calculated by applying the `compare_conditions()` function to an `opafit` object produced by a call to `opa()`:

```{r}
condition_comparisons <- compare_conditions(opamod)

condition_comparisons$pccs
condition_comparisons$cvals
```

### Multiple groups 

If the data consist of multiple groups a categorical grouping variable can be passed with the `group` keyword to produce results for each group within the data, in addition to individual results.

```{r group_data}
dat$group <- rep(c("A", "B", "C", "D"), 5)
dat$group <- factor(dat$group, levels = c("A", "B", "C", "D"))

opamod2 <- opa(dat[, 1:4], h, group = dat$group, cval_method = "exact")
```

The summary output displays results organised by group.

```{r summary_opamod2}
summary(opamod2, digits = 3)
```

Similarly, plotting the output shows individual PCCs and c-values by group.

```{r plot_opamod2, fig.width=8, fig.height=5, fig.align="center"}
plot(opamod2)
```

## Acknowledgements

Development of `opa` was supported by a [Medical Research Foundation](https://www.medicalresearchfoundation.org.uk/) Fellowship (MRF-049-0004-F-BEEC-C0899).
