---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)

# set plot colors
palette(c("#0073C2", "#EFC000", "#868686", "#CD534C"))
```

# opa <a href="https://timbeechey.github.io/opa/"></a>

<!-- badges: start -->
![](https://www.r-pkg.org/badges/version-ago/opa?color=orange)
![](https://cranlogs.r-pkg.org/badges/grand-total/opa) 
[![R-CMD-check](https://github.com/timbeechey/opa/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/timbeechey/opa/actions/workflows/R-CMD-check.yaml)
[![codecov](https://codecov.io/gh/timbeechey/opa/graph/badge.svg?token=Q3ZI7BBMIK)](https://codecov.io/gh/timbeechey/opa)
<!-- badges: end -->

An R package for ordinal pattern analysis.

## Installation

opa can be installed from CRAN with:

```{r install_opa, eval=FALSE}
install.packages("opa")
```

You can install the development version of opa from [GitHub](https://github.com/) with:

```{r install_github, eval=FALSE}
# install.packages("remotes")
remotes::install_github("timbeechey/opa")
```

## Using `opa`

See the [introductory guide](https://timbeechey.github.io/opa/articles/opa.html) for a brief demonstration of how to conduct an ordinal pattern analysis using `opa`.

## Acknowledgements

Development of `opa` was supported by a [Medical Research Foundation](https://www.medicalresearchfoundation.org.uk/) Fellowship (MRF-049-0004-F-BEEC-C0899).
