---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)

# set plot colors
palette("Set 2")
#palette(adjustcolor(palette(), alpha.f = 0.7))
```

# opa

<!-- badges: start -->
![](https://www.r-pkg.org/badges/version-ago/opa?color=orange)
![](https://cranlogs.r-pkg.org/badges/grand-total/opa) 
<!-- badges: end -->

An R package for ordinal pattern analysis.

## Installation

opa can be installed from CRAN with:

```{r install_opa, eval=FALSE}
install.packages("opa")
```

You can install the development version of opa from [GitHub](https://github.com/) with:

```{r install_github, eval=FALSE}
# install.packages("devtools")
devtools::install_github("timbeechey/opa")
```

## Citation

To cite opa in your work you can use the output of:

```{r citation, eval=FALSE, message=FALSE, warning=FALSE}
citation(package = "opa")
```

## How ordinal pattern analysis works

Ordinal pattern analysis is similar to Kendall's Tau. Whereas Kendall's tau is a measure of similarity between two data sets in terms of rank ordering, ordinal pattern analysis is intended to quantify the match between an hypothesis and patterns of individual-level data across conditions or mesaurement instances.

Ordinal pattern analysis works by comparing the relative ordering of pairs of observations and computing whether these pairwise relations are matched by an hypothesis. Each pairwise ordered relation is classified as an increases, a decrease, or as no change. These classifications are encoded as 1, -1 and 0, respectively. An hypothesis of a monotonic increase in the response variable across four experimental conditions can be specified as:

```{r hypothesis1}
h <- c(1, 2, 3, 4)
```

The hypothesis `h` encodes six pairwise relations, all increases: `1 1 1 1 1 1`.

A row of individual data representing measurements across four conditions, such as:

```{r vector1}
dat <- c(65.3, 68.8, 67.0, 73.1)
```

encodes the ordered pairwise relations `1 1 1 -1 1 1`. The percentage of orderings which are correctly classified by the hypothesis (PCC) is the main quantity of interest in ordinal pattern analysis. Comparing `h` and `dat`, the PCC is `5/6 = 0.833` or 83.3%. An hypothesis which generates a greater PCC is preferred over an hypothesis which generates a lower PCC for given data.

It is also possible to calculate a chance-value for a PCC which is equal to the chance that a PCC at least as great as the PCC of the observed data could occur as a result of a random ordering of the data. Chance values can be computed using either a permutation test or a randomization test.

## Using `opa`

```{r load_opa}
library(opa)
```

A hypothesized relative ordering of the response variable across conditions is specified with a numeric vector:

```{r hypothesis2}
h <- c(1, 2, 4, 3)
```

The hypothesis can be visualized with the `plot_hypothesis()` function:

```{r plot_hypothesis, fig.width=4, fig.height=4, fig.align="center"}
plot_hypothesis(h)
```

Data should be in _wide_ format with one column per measurement condition and one row per individual:

```{r load_data}
set.seed(123)

dat <- data.frame(t1 = rnorm(20, mean = 12, sd = 2),
                  t2 = rnorm(20, mean = 15, sd = 2),
                  t3 = rnorm(20, mean = 20, sd = 2),
                  t4 = rnorm(20, mean = 17, sd = 2))
                  
round(dat, 2)
```

An ordinal pattern analysis model to consider how the hypothesis `h` matches each individual pattern of results in `dat` can be fitted using:

```{r opamod1}
opamod <- opa(dat, h, cval_method = "exact")
```

A summary of the model output can be viewed using:

```{r summary1}
summary(opamod)
```

Individual-level model output can be visualized using:

```{r plot_opamod1, fig.width=9, fig.height=4.5, fig.align="center"}
plot(opamod)
```

To aid in visual interpretation, iIndividual PCCs and c-values can also be plotted relative to specified thresholds:

```{r threshold plots, fig.width=5, fig.height=5, fig.align="center", fig.show="hold", out.width="50%"}
pcc_plot(opamod, threshold = 90)
cval_plot(opamod, threshold = 0.1)
```


### Pairwise comparison of measurement conditions

Pairwise comparisons of measurement conditions can be calculated by applying the `compare_conditions()` function to an `opafit` object produced by a call to `opa()`:

```{r}
condition_comparisons <- compare_conditions(opamod)

condition_comparisons$pccs
condition_comparisons$cvals
```

### Multiple groups 

If the data consist of multiple groups a categorical grouping variable can be passed with the `group` keyword to produce results for each group within the data, in addition to individual results.

```{r group_data}
dat$group <- rep(c("A", "B", "C", "D"), 5)
dat$group <- factor(dat$group, levels = c("A", "B", "C", "D"))

opamod2 <- opa(dat[, 1:4], h, group = dat$group, cval_method = "exact")
```

The summary output displays results organised by group.

```{r summary_opamod2}
summary(opamod2)
```

Similarly, plotting the output shows individual PCCs and c-values by group.

```{r plot_opamod2, fig.width=10, fig.height=6, fig.align="center"}
plot(opamod2)
```

## Acknowledgements

Development of `opa` was supported by a [Medical Research Foundation](https://www.medicalresearchfoundation.org.uk/) Fellowship (MRF-049-0004-F-BEEC-C0899).
