---
title: "opa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{opa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Background

`opa` is an implementation of methods described in publications including [Thorngate (1987)](https://doi.org/10.1016/S0166-4115(08)60083-7) and [Grice et al. (2015)](https://doi.org/10.1177/2158244015604192). Thorngate (1987) attributes the original idea to:

Parsons, D. (1975). _The directory of tunes and musical themes_. S. Brown.

## How ordinal pattern analysis works

Ordinal pattern analysis is a simple non-parametric method, similar to Kendall's Tau. Whereas Kendall's tau is a measure of similarity between two data sets in terms of rank ordering, ordinal pattern analysis is intended to quantify the match between a hypothesis and patterns of individual-level data across conditions or measurement instances.

Ordinal pattern analysis works by comparing the relative ordering of pairs of observations and computing whether those pairwise relations are matched by a hypothesis. Each pairwise ordered relation is classified as an increase, a decrease, or as no change. These classifications are encoded as 1, -1 and 0, respectively. For example, a hypothesis of a monotonic increase in a response variable across four experimental conditions can be specified as:

```{r hypothesis1}
(h <- c(1, 2, 3, 4))
```

Note that the absolute values are not important, only their relative ordering. The hypothesis `h` encodes six pairwise relations, all increases: `1 1 1 1 1 1`.

A row of individual data representing measurements across four conditions, such as:

```{r vector1}
dat <- c(65.3, 68.8, 67.0, 73.1)
```

encodes six ordered pairwise relations `1 1 1 -1 1 1`. The percentage of orderings which are correctly classified by the hypothesis (_PCC_) is the main quantity of interest in ordinal pattern analysis. Comparing `h` and `dat`, the PCC is `5/6 = 0.833` or 83.3%. A hypothesis which generates a greater PCC is preferred over a hypothesis which generates a lower PCC for given data.

It is also possible to calculate a _chance-value_ for a PCC which is equal to the chance that a PCC at least as great as the PCC of the observed data could occur as a result of a random re-ordering of the data. Chance values can be computed using a randomization test.

## Modeling repeated measures data

Once installed, you can load `opa` with

```{r load_opa}
library(opa)
```

### Data

For this example we will use the `sleepstudy` data from the `lme4` package.

```{r load_data}

```

#### Wide format

## Specifying a hypothesis

## Fitting an ordinal pattern analysis model

## References

Grice, J. W., Craig, D. P. A., & Abramson, C. I. (2015). A Simple and Transparent Alternative to Repeated Measures ANOVA. SAGE Open, 5(3), 215824401560419. https://doi.org/10.1177/2158244015604192

Parsons, D. (1975). _The directory of tunes and musical themes_. S. Brown.

Thorngate, W. (1987). Ordinal Pattern Analysis: A Method for Assessing Theory-Data Fit. Advances in Psychology, 40, 345–364. https://doi.org/10.1016/S0166-4115(08)60083-7


